Gait Segmentation Model for Rotation Time Series Collected by a Wearable Sensor using Machine Learning

### 2025-03-31 article_segmentation.qmd
Tree models, data splitting and tuning strategy.

Finally, the variable to be predicted contains the two possible classes *Hell-Strike* and *other times* from the GAITRite© mat output. We choose to label a number of points as *Hell-Strike* around the precise event time. This allows to take into account a some uncertainty, as the time range between two points is only 10 ms. By labelling as *Hell-Strike* seven points rather than just one, the event happens in a window of 70 ms. This strategy also reduces the imbalance between the two classes.

@@

@@

## Data splitting

## Tuning strategy

@@

>>>>

Finally, the variable to be predicted contains the two possible classes *Heel-Strike* and *other times* from the GAITRite© mat output. We choose to label a number of points as *Heel-Strike* around the precise event time. This allows to take into account a some uncertainty, as the time range between two points is only 10 ms. By labelling as *Heel-Strike* seven points rather than just one, the event happens in a window of 70 ms. This strategy also reduces the imbalance between the two classes.

@@

Decision tree

: Decision trees consist of a nested sequence of if-then statements for the predictor classifying data. Observations are assigned to their class according to the variable values. Here is a figure of a tree with two splitting, leading to three leaf nodes.

: ![Basic decision tree.](images/decision-tree.png){#fig-decisiontree width=250}

: The goal is to classify the observations into smaller and more homogenous groups, forming rectangular areas within data points. Homogeneity is defined here as how pure are the splitting nodes [@kuhn2013applied], meaning that there is a higher proportion of one class in each node. We can use the Gini index [@breiman2017classification] to compute purity. For a binary classification, if $p1$ and $p2$ are the probabilities for class 1 and 2 respectively, the Gini index for a node computed with the formula:

: $$
\text{Gini} = p_1(1-p_1) + p_2(1-p_2) = 2p_1p_2
$${#eq-gini}

: This index is minimal when one of the probabilities tends toward zero, indicating that the node is pure. To build the tree, the model builds splitting nodes by evaluating each splitting point (ie values taken by variables to split observations into groups) to find the one minimizing the Gini index, until a stopping criteria is met.

Bagged trees and Random forest

: A bagged trees model is an ensemble of $M$ decision trees. Each decision tree is built with a boostrap sample of the original data, which is a sample of same size than the original data by selecting random observations with replacement. Then, each of these trees is used to generate a prediction for a new observation. Finally, the observation is classified in the class that has collected the greatest number of votes from the trees.

: A random forest is a quite similar model, also using a number of decision trees. The model also train a tree on a boostrap sample of the original data. The difference is that, at each split, a random subset of the variables is selected to then find the best predictors within this subset. The observations are then classified as usual. The prediction is done the same way as seen before, with each tree voting for a class and selecting the majority to classify the observation.

Neural networks

: *to do*

@@

## Dealing with imbalanced data

By classifiying specific times in a time serie, we work with imbalanced data, meaning that a small proportion of our data is in the target class, and most of the data points are in the other class. In this situation, the model learns adequate informations about the majority class but doesn't have enough information on the minority class. This implies bad predictions on the target class because the model misses this class.

A lot of sampling algorithms exist to adress this issue such as random oversampling or random undersampling which add or remove observations from certain classes to create a balanced dataset. More complex methods exist, we can cite the popular SMOTE [@chawla2002] and ADASYN [@haibo2008] algorithms that create synthetic data by considering the classes of the nearest neighbors of random observations. Although these algorithms are very popular, they did not show good results for our data. We are making the hypothesis that our data is too imbalanced for these methods.

To tell the model to pay more attention to the patterns in the minority class, an effective way is to put a larger weight on this class. More precisely, the weight given to each observation specifies how much each observation influences the model's estimation In order for the model to be biased in favor of the observations considered more important, in our case those belonging to the minority class, the weights of the observations are integrated into the cost function. This makes it possible to regulate the cost of misclassification, in the sense that misclassifying more important observations will be more costly, encouraging the model to avoid this situation [@hashemi2018].

Weights put on classes are the inverse class frequency weights. The weight on the class $j$ is the following:

$$
\omega_j = \frac{n}{n_{classes} \hspace{1mm} n_j}
$$ {#eq-inv-class-frequency}

with:

- $n$ the total number of observations.
- $n_{classes}$ the number of classes.
- $n_j$ the number of observations in the class $j$.

This formula gives balanced weights at the different classes to pay better attention at the minority class.

@@

In the case of imbalanced class, we cannot use accucary to correctly evaluate the model. Indeed, since the model will predict almost only the majority class, the predictions in this class will be good, implying that the accuracy will be high because most of the observations will be well-classified. This does not take into account the observations in the minority class which are not predicted.

We need to use other metrics that are adapted for this situation. An important tool is the confusion matrix which indicate the number of predictions in each class compared to the real values of the observations.

*Here, either metric for binary or multiclass*

## Data splitting and Tuning strategy

To correctly tune and evaluate the model, data is split in three sets (see @fig-datasplitting).

![Data splitting.](images/data-splitting.png){#fig-datasplitting width=300}

During tuning and training, the training set is used to train the model and the validation set is used to find the best parameters. The test set is never used during the training step to then evaluate the model on data never seen before.

More precisely, for a neural network, we first use the training set to train the parameters (the weights and the biases of the network). Then, we use the validation set to determine the hyper-parameters (for instance, the number of layers and nodes). Finally, we use the test set to evaluate the model on unseen data.

To tune hyper-parameters, automated tuning was used by grid search which is an exhaustive search in the hyper-parameter space. For each of our parameter, we set a grid of values to be tested. Every combination of the values is then evaluated to get the best possible combination of hyper-parameter values.
